$\pi$ with leftovers: \\ a mechanisation in Agda

### 2021-01-20 paper/paper.tex
Bring theorems back into the meta theory section

Leftover typing also allows us to prove a \emph{framing} theorem, which states that arbitrary usage annotations can be added to the input usage context, as long as they are added to the output usage context as well -- it is a way of understanding a typing derivation independently from its unused resources.
An example use of this is showing that $\constr{comp-comm}$ preserves typing: we use framing to show that in $\comp{P}{Q}$ the typing of $P$ and $Q$ is independent of one another.
Moreover, leftover typing allows \emph{weakening} and \emph{strengthening} to acquire a more general form, in which variables can freely be added or removed from context regardless of their usage annotation -- as long as this usage annotation is added or removed from the leftover context as well.

>>>>

Leftover typing also allows us to prove a \emph{framing} theorem, which states that arbitrary usage annotations can be added to the input usage context, as long as they are added to the output usage context as well -- one can thus understand a typing derivation independently from its unused resources.
An example use of this is showing that $\constr{comp-comm}$ preserves typing: we use framing to show that in $\comp{P}{Q}$ the typing of $P$ and $Q$ is independent of one another.
\begin{nitheorem}[Framing]
  \label{thm:framing}
  Let $P$ be well typed in $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Let $\Delta$ be such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Let $\Gamma_r$ and $\Xi_r$ be arbitrary contexts such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}

Leftover typing allows \emph{weakening} and \emph{strengthening} to acquire a more general form, in which variables can freely be added or removed from context regardless of their usage annotation -- as long as this usage annotation is added or removed from the leftover context as well.
Weakening states that inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.
Strengthening states that removing an unused variable preserves the well-typedness of a process.

\begin{nitheorem}[Weakening]
  \label{thm:weakening}
  Let $\func{ins}_i$ insert an element into a context at position $i$ --- for simplicity, we use it both to insert types into typing contexts and usage annotations into usage contexts.
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, lifting every variable greater than or equal to $i$ in $P$ is well typed in
  $\types{\func{ins}_i \; t \; \gamma}{\func{ins}_i \; x \; \Gamma}{\func{lift}_i \; P}{\func{ins}_i \; x \; \Xi}$.
\end{nitheorem}

\begin{nitheorem}[Strengthening]
  \label{thm:strengthening} 
  Let $\func{del}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable not in $P$, such that $uP \; : \; \Unused_i \; P$.
  Then lowering every variable greater than $i$ in $P$ is well typed in $\types{\func{del}_i \; \gamma}{\func{del}_i \; \Gamma}{\func{lower}_i \; P \; uP}{\func{del}_i \; \Xi}$.
\end{nitheorem}

\subsubsection{Type Safety}
Subject congruence states that applying structural congruence (\autoref{structural-congruence}) to a well-typed process preserves its well-typedness.
Subject reduction states that a well-typed process that communicates on a channel $c$ (\autoref{operational-semantics}) preserves its well-typedness and consumes $\lio$ from $c$ if $c$ is a channel in the environment.

\begin{nitheorem}[Subject congruence]
  \label{thm:subject-congruence}
  If $P \eq{} Q$ and $\types{\gamma}{\Gamma}{P}{\Xi}$, then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{nitheorem}

\begin{nitheorem}[Subject reduction]
  \label{thm:subject-reduction}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$ and reduce such that $P \reduce{c} Q$.
  \begin{itemize}
    \item If $c$ is $\constr{internal}$, then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
    \item If $c$ is $\constr{external} \; i$ and $\containsusage{\Gamma}{i}{\lio}{\Delta}$, then $\types{\gamma}{\Delta}{Q}{\Xi}$.
  \end{itemize}
\end{nitheorem}
